---
title: "What can the character state database tell us?"
subtitle: "Preliminary overview of the Known Hominin Characters database"
author: "Jyhreh Johnson"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,          # code is folded in the HTML; set FALSE to hide entirely
                      message = FALSE, warning = FALSE,
                      fig.width = 7.5, fig.height = 4.5, dpi = 150)

library(tidyverse)   # dplyr, ggplot2, forcats, stringr, tidyr, readr
library(knitr)
library(scales)
```

```{r helpers}
# ---- Settings you may want to change ---------------------------------------
data_file <- "Known_Hominin_Characters.csv"

# Spreadsheet codes -> readable labels. Anything not listed here is kept as-is
# (so a new code added to the spreadsheet shows up instead of vanishing).
region_lookup      <- c(CR = "Cranial", DEN = "Dental", CF = "Craniofacial", PC = "Postcranial")
chartype_lookup    <- c(C = "Continuous", D = "Discrete")
multistate_lookup  <- c(OM = "Ordered", UM = "Unordered")
designation_lookup <- c(PHY = "Phylogenetic", TAX = "Taxonomic", BOTH = "Both")
signif_lookup      <- c(D = "D", ND = "ND")   # confirm what D / ND stand for and relabel here

# How to tell quantitative from qualitative DISCRETE characters. The database has no field for this,
# so it is inferred from the character name (edit the patterns, or list ids to override):
measure_pattern <- "ratio|index|length|breadth|width|size|height|depth|thickness"   # measurement-based (binned) traits
meristic_pattern <- "\\bnumber\\b|\\bcount\\b"                                        # counts, e.g. "Cusp number on M1"
force_quantitative_ids <- c()     # ids to always treat as quantitative, e.g. c(12, 57)
force_qualitative_ids  <- c()     # ids to always treat as qualitative (removes false positives)

fig_dir     <- "figures"          # every plot is saved here
fig_formats <- c("png", "pdf")    # add "svg" or "tiff" if a journal/slide template needs it
fig_dpi     <- 300
# ----------------------------------------------------------------------------

# ---- Figure saving ----------------------------------------------------------
# show_plot() saves a plot to fig_dir in each format in fig_formats, remembers it for the
# combined PDF at the end, and prints it (so it appears in the report and in the RStudio
# Plots pane). Usage:  p <- ggplot(...);  show_plot(p, "my_figure_name", height = 5)
dir.create(fig_dir, showWarnings = FALSE, recursive = TRUE)
.figs <- new.env()
.figs$plots <- list()
show_plot <- function(p, name, width = 7.5, height = 4.5) {
  for (fmt in fig_formats) {
    ggsave(file.path(fig_dir, paste0(name, ".", fmt)), plot = p,
           width = width, height = height, dpi = fig_dpi, bg = "white")
  }
  .figs$plots[[name]] <- p
  print(p)
  invisible(p)
}

# Recode a spreadsheet column using a lookup; blanks/NA become "Unscored".
# Every level in the lookup is kept even if it has 0 rows (e.g. Postcranial),
# so gaps in the database are visible in tables and plots.
recode_col <- function(x, lookup, na_label = "Unscored", keep_all_levels = TRUE) {
  x0  <- str_to_upper(str_trim(as.character(x)))
  x0[x0 == ""] <- NA
  # accept either the spreadsheet code (CR) or the full label (Cranial), in any capitalization
  lookup_all <- c(lookup, setNames(unname(lookup), str_to_upper(unname(lookup))))
  out <- unname(lookup_all[x0])
  out <- ifelse(is.na(x0), na_label, ifelse(is.na(out), x0, out))   # unknown codes pass through
  lev <- if (keep_all_levels) unname(lookup) else intersect(unname(lookup), out)
  lev <- c(lev, setdiff(sort(unique(out)), c(lev, na_label)), if (any(is.na(x0))) na_label)
  factor(out, levels = lev)
}

# Count + proportion table for one or more grouping columns
count_prop <- function(df, ...) {
  df %>%
    count(..., name = "n", .drop = FALSE) %>%   # .drop = FALSE keeps empty factor levels
    mutate(prop = n / sum(n),
           pct  = percent(prop, accuracy = 0.1))
}

# Single-variable bar plot with "n (pct)" labels
bar_plot <- function(tab, x, title, xlab, fill = "#4C78A8") {
  ggplot(tab, aes(x = {{ x }}, y = n)) +
    geom_col(fill = fill, width = 0.7) +
    geom_text(aes(label = paste0(n, "\n(", pct, ")")), vjust = -0.3, size = 4) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
    labs(title = title, x = xlab, y = "Number of characters") +
    theme_minimal(base_size = 14) +
    theme(panel.grid.major.x = element_blank())
}

# 100%-stacked bar: composition of `fill` within each level of `x`, labeled with counts
stack_plot <- function(tab, x, fill, title, xlab, filllab, palette = "Set2") {
  tab %>%
    group_by({{ x }}) %>% mutate(share = n / sum(n)) %>% ungroup() %>%
    ggplot(aes(x = {{ x }}, y = share, fill = {{ fill }})) +
    geom_col(width = 0.65) +
    geom_text(aes(label = ifelse(n > 0, n, "")), position = position_stack(vjust = 0.5), size = 4) +
    scale_y_continuous(labels = percent, expand = expansion(mult = c(0, 0.02))) +
    scale_fill_brewer(palette = palette, name = filllab, drop = FALSE) +
    labs(title = title, x = xlab, y = "Share of characters") +
    theme_minimal(base_size = 14)
}

# Add a "Sum" row and column to a 2-way table (base addmargins() errors on some R versions)
add_totals <- function(xt) {
  m <- matrix(as.numeric(xt), nrow = nrow(xt), dimnames = dimnames(xt))
  rbind(cbind(m, Sum = rowSums(m)), Sum = c(colSums(m), sum(m)))
}

# Monte Carlo chi-square + Cramer's V for a contingency table (zero rows/cols dropped)
assoc_test <- function(xt, B = 10000, seed = 1) {
  xt <- xt[rowSums(xt) > 0, colSums(xt) > 0, drop = FALSE]
  set.seed(seed)
  chi <- chisq.test(xt, simulate.p.value = TRUE, B = B)
  v <- sqrt(unname(chi$statistic) / (sum(xt) * (min(dim(xt)) - 1)))
  list(chi = chi, v = v, table = xt)
}
```

# Load data

```{r load}
# readr handles the BOM automatically and tolerates rows with a stray trailing comma
# (read.csv fails on those). Everything is read as text, then cleaned below.
cat("Reading:", normalizePath(data_file, mustWork = FALSE), "\n")
cat("Working directory:", getwd(), "\n")
hominin_characters <- readr::read_csv(data_file, col_types = readr::cols(.default = "c"),
                                      name_repair = "minimal", show_col_types = FALSE) %>%
  as.data.frame(check.names = FALSE)
# Clean column names: remove BOM / non-breaking spaces and collapse repeated spaces
names(hominin_characters) <- names(hominin_characters) %>%
  str_replace_all("[\uFEFF\u00A0]", " ") %>% str_squish()
hominin_characters$id <- as.integer(hominin_characters$id)

needed <- c("id", "character", "skeletal_element", "anatomical_region", "character_type",
            "multistate_type", "character_designation", "character_significance",
            "uberon_ID", "number_of_states")
missing_cols <- setdiff(needed, names(hominin_characters))
if (length(missing_cols) > 0) {
  stop("These columns were not found in ", data_file, ": ", paste(missing_cols, collapse = ", "),
       "\nColumns present: ", paste(names(hominin_characters), collapse = ", "))
}

hc <- hominin_characters %>%
  mutate(
    anatomical_region      = recode_col(anatomical_region, region_lookup),
    character_type         = recode_col(character_type, chartype_lookup),
    multistate_type        = recode_col(multistate_type, multistate_lookup),
    character_designation  = recode_col(character_designation, designation_lookup),
    character_significance = recode_col(character_significance, signif_lookup,
                                        na_label = "Not assessed", keep_all_levels = FALSE),
    skeletal_element       = str_squish(skeletal_element),            # fixes "Face" vs "Face "
    skeletal_element       = replace_na(na_if(skeletal_element, ""), "Unscored"),
    uberon_ID              = na_if(str_trim(uberon_ID), ""),
    number_of_states       = suppressWarnings(as.numeric(number_of_states))
  )

n_chars <- nrow(hc)

# Classify each character as continuous, discrete-quantitative, or discrete-qualitative.
# Discrete-quantitative = discrete states that encode a count or a binned measurement.
hc <- hc %>%
  mutate(
    quant_reason = case_when(
      id %in% force_qualitative_ids  ~ NA_character_,
      id %in% force_quantitative_ids ~ "Manual override",
      str_detect(character, regex(meristic_pattern, ignore_case = TRUE)) ~ "Count (meristic)",
      str_detect(character, regex(measure_pattern,  ignore_case = TRUE)) ~ "Measurement-based",
      TRUE ~ NA_character_),
    trait_class = case_when(
      character_type == "Continuous"                    ~ "Continuous (quantitative)",
      character_type == "Discrete" & !is.na(quant_reason) ~ "Discrete, quantitative",
      character_type == "Discrete"                      ~ "Discrete, qualitative",
      TRUE                                              ~ "Unscored"),
    trait_class = factor(trait_class, levels = c("Continuous (quantitative)", "Discrete, quantitative",
                                                 "Discrete, qualitative", "Unscored")))

# Locate the four source-text columns by keyword, so small differences in the header
# (spacing, punctuation) don't break the script
find_col <- function(df, pattern) {
  hit <- names(df)[str_detect(names(df), regex(pattern, ignore_case = TRUE))]
  if (length(hit) > 0) hit[1] else NA_character_
}
source_cols <- c(collard = find_col(hc, "collard"),
                 dembo   = find_col(hc, "dembo"),
                 strait_1997        = find_col(hc, "strait.*1997"),
                 strait_grine_2004  = find_col(hc, "strait.*grine"))
source_cols <- source_cols[!is.na(source_cols)]
dembo_col   <- if ("dembo" %in% names(source_cols)) source_cols[["dembo"]] else NA_character_

cat("Columns in the CSV:", paste(names(hominin_characters), collapse = " | "), "\n")
cat("Source columns recognized:", paste(source_cols, collapse = " | "), "\n")

# Diagnostic: what values did the script find? If a table below looks wrong, the spreadsheet
# codes probably differ from the lookups at the top of the script.
for (col in c("character_type", "anatomical_region", "multistate_type", "character_designation")) {
  cat("\n", col, ":\n", sep = ""); print(table(hc[[col]], useNA = "ifany"))
}
if (!any(hc$character_type %in% c("Continuous", "Discrete"))) {
  stop("No character_type values matched 'Continuous' or 'Discrete'. Values found: ",
       paste(unique(hominin_characters$character_type), collapse = ", "),
       ". Edit chartype_lookup at the top of the script to match.")
}
```

The database currently contains **`r n_chars`** characters (IDs `r min(hc$id)`-`r max(hc$id)`).

# Q1. How many characters are continuous vs. discrete?

Continuous characters are quantitative (measurements). Discrete characters are scored as states, but some of those states encode a *count* (for example "Cusp number on M1") or a *binned measurement* (for example "Upper facial breadth"), so the discrete bar is split into quantitative and qualitative parts.

```{r q1}
q1 <- count_prop(hc, character_type)
kable(q1, col.names = c("Character type", "n", "Proportion", "%"), digits = 3,
      caption = "Table 1. Character type counts")

# Split of the discrete characters
q1b <- count_prop(hc, trait_class) %>% filter(n > 0 | trait_class != "Unscored")
kable(q1b, col.names = c("Trait class", "n", "Proportion of all characters", "%"), digits = 3,
      caption = "Table 1b. Continuous vs. discrete-quantitative vs. discrete-qualitative")

hc %>% filter(character_type == "Discrete") %>%
  count(`Basis for 'quantitative'` = replace_na(quant_reason, "None (qualitative)"), name = "n") %>%
  kable(caption = "Table 1c. Why discrete characters were classed as quantitative")

# Stacked bar: the Discrete bar is split into quantitative and qualitative segments
q1_plot_data <- hc %>%
  mutate(bar = case_when(character_type == "Continuous" ~ "Continuous",
                         character_type == "Discrete"   ~ "Discrete",
                         TRUE                           ~ "Unscored"),
         bar = factor(bar, levels = c("Continuous", "Discrete", "Unscored"))) %>%
  count(bar, trait_class) %>%
  mutate(label = ifelse(bar == "Discrete", paste0(n, " (", percent(n / n_chars, accuracy = 0.1), ")"), ""))  # segment labels only where a bar is split
q1_totals <- q1_plot_data %>% group_by(bar) %>% summarise(total = sum(n), .groups = "drop") %>%
  mutate(label = paste0("n = ", total, " (", percent(total / n_chars, accuracy = 0.1), ")"))

trait_colors <- c("Continuous (quantitative)" = "#E78AC3",
                  "Discrete, quantitative"    = "#2B6CB0",
                  "Discrete, qualitative"     = "#9ECAE1",
                  "Unscored"                  = "#BDBDBD")

p_q1 <- ggplot(q1_plot_data, aes(x = bar, y = n, fill = trait_class)) +
  geom_col(width = 0.65, colour = "white") +
  geom_text(aes(label = label, colour = trait_class), position = position_stack(vjust = 0.5), size = 3.8,
            show.legend = FALSE) +
  geom_text(data = q1_totals, aes(x = bar, y = total, label = label),
            vjust = -0.5, size = 4.2, fontface = "bold", inherit.aes = FALSE) +
  scale_fill_manual(values = trait_colors, name = NULL, breaks = names(trait_colors)[1:3]) +
  scale_colour_manual(values = c("Continuous (quantitative)" = "black", "Discrete, quantitative" = "white",
                                 "Discrete, qualitative" = "black", "Unscored" = "black")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(title = "Continuous vs. discrete characters", x = "Character type", y = "Number of characters") +
  theme_minimal(base_size = 14) +
  theme(panel.grid.major.x = element_blank(), legend.position = "bottom")
show_plot(p_q1, "q1_character_type")

# Review list: check these by eye, then use force_quantitative_ids / force_qualitative_ids at the top
write.csv(hc %>% filter(character_type == "Discrete", !is.na(quant_reason)) %>%
            select(id, character, skeletal_element, number_of_states, quant_reason),
          "discrete_quantitative_review.csv", row.names = FALSE)
```

`r sum(hc$trait_class == "Discrete, quantitative")` of the `r sum(hc$character_type == "Discrete")` discrete characters (`r percent(mean(hc$trait_class[hc$character_type == "Discrete"] == "Discrete, quantitative"), accuracy = 1)`) look quantitative, so the database is `r percent(mean(hc$trait_class %in% c("Continuous (quantitative)", "Discrete, quantitative")), accuracy = 1)` quantitative overall. This split is a name-based screen, because no field in the database records it. Review `discrete_quantitative_review.csv` and correct any misclassified characters with `force_quantitative_ids` / `force_qualitative_ids` in the settings at the top. Adding a quantitative/qualitative column to the spreadsheet would make this exact.

# Q2. How many characters are cranial, dental, craniofacial, or postcranial?

Regions with no characters are kept (n = 0), so a gap in the database is visible. **Postcranial is currently empty.**

```{r q2}
q2 <- count_prop(hc, anatomical_region)
kable(q2, col.names = c("Anatomical region", "n", "Proportion", "%"), digits = 3,
      caption = "Table 2. Characters by anatomical region")

p_q2 <- bar_plot(q2, anatomical_region, "Characters by anatomical region", "Anatomical region", "#66C2A5")
show_plot(p_q2, "q2_anatomical_region")
```

# Q3. Do continuous and discrete characters sample the skeleton differently?

Only characters with both a region and a type are used here.

```{r q3-table}
q3 <- hc %>%
  filter(character_type %in% c("Continuous", "Discrete"),
         !anatomical_region %in% c("Unscored", "Postcranial")) %>%
  droplevels() %>%
  count_prop(anatomical_region, character_type)

if (nrow(q3) == 0) stop("Q3 has no rows: no character has both a recognized region and a recognized type. ",
                        "Check the value tables printed in the 'load' chunk.")
xt <- xtabs(n ~ anatomical_region + character_type, data = q3)
kable(add_totals(xt), caption = "Table 3a. Region x character type (counts)")

kable(round(100 * prop.table(xt, margin = 2), 1),
      caption = "Table 3b. Region composition within each character type (% of column)")
```

```{r q3-plot}
p_q3 <- stack_plot(q3, character_type, anatomical_region,
                   "Anatomical region by character type", "Character type", "Region")
show_plot(p_q3, "q3_region_by_character_type")
```

```{r q3-test}
a3 <- assoc_test(xt)
a3$chi
kable(round(a3$chi$stdres, 2),
      caption = "Table 3c. Standardized residuals (|value| > 2 = cell differs from expectation)")
```

Chi-square (Monte Carlo) p = `r signif(a3$chi$p.value, 3)`, Cramér's V = `r round(a3$v, 2)`. Report the effect size with the p-value. Note that continuous characters are few (`r sum(q3$n[q3$character_type == "Continuous"])`), so cell counts for them are small; treat the test as descriptive.

Optional: the `ggstatsplot` version of the same figure (uncomment to run).

```{r q3-ggstatsplot, eval=FALSE}
library(ggstatsplot)
ggbarstats(data = hc %>% filter(character_type %in% c("Continuous", "Discrete")),
           x = anatomical_region, y = character_type,
           title = "Anatomical region by character type")
```

# Q4. How many characters are ordered vs. unordered?

In this database *continuous* characters also carry an ordered/unordered designation (they are binned into states), so this is shown for all characters, split by type.

```{r q4}
q4 <- count_prop(hc, multistate_type)
kable(q4, col.names = c("Multistate type", "n", "Proportion", "%"), digits = 3,
      caption = "Table 4. Ordered vs. unordered characters")

q4b <- hc %>% filter(character_type != "Unscored", multistate_type != "Unscored") %>%
  droplevels() %>% count_prop(character_type, multistate_type)
p_q4b <- stack_plot(q4b, character_type, multistate_type,
                    "Ordered vs. unordered, by character type", "Character type", "Multistate type", "Set3")
show_plot(p_q4b, "q4_ordering_by_character_type")
```

```{r q4-region}
q4c <- hc %>% filter(anatomical_region %in% c("Cranial", "Dental", "Craniofacial"),
                     multistate_type != "Unscored") %>%
  droplevels() %>% count_prop(anatomical_region, multistate_type)
p_q4c <- stack_plot(q4c, anatomical_region, multistate_type,
                    "Ordered vs. unordered, by region", "Anatomical region", "Multistate type", "Set3")
show_plot(p_q4c, "q4_ordering_by_region")

a4 <- assoc_test(xtabs(n ~ anatomical_region + multistate_type, data = q4c))
```

Region x ordering: chi-square (Monte Carlo) p = `r signif(a4$chi$p.value, 3)`, Cramér's V = `r round(a4$v, 2)`.

# Q5. How many states do characters have?

```{r q5}
q5 <- hc %>% filter(!is.na(number_of_states)) %>%
  count(number_of_states, name = "n") %>%
  mutate(pct = percent(n / sum(n), accuracy = 0.1))
kable(q5, col.names = c("Number of states", "n", "%"), caption = "Table 5. Number of states per character")

p_q5 <- ggplot(q5, aes(x = factor(number_of_states), y = n)) +
  geom_col(fill = "#A6D854", width = 0.7) +
  geom_text(aes(label = n), vjust = -0.4) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(title = "Number of character states", x = "Number of states", y = "Number of characters") +
  theme_minimal(base_size = 14) + theme(panel.grid.major.x = element_blank())
show_plot(p_q5, "q5_number_of_states")
```

`r sum(hc$number_of_states == 2, na.rm = TRUE)` characters (`r percent(mean(hc$number_of_states == 2, na.rm = TRUE), accuracy = 1)` of those with a state count) are binary.

# Q6. Which skeletal elements are most heavily sampled?

```{r q6, fig.height = 6}
q6 <- hc %>% count(skeletal_element, anatomical_region, name = "n") %>%
  group_by(skeletal_element) %>% mutate(total = sum(n)) %>% ungroup()

p_q6 <- q6 %>%
  mutate(skeletal_element = fct_reorder(skeletal_element, total)) %>%
  ggplot(aes(x = skeletal_element, y = n, fill = anatomical_region)) +
  geom_col() +
  coord_flip() +
  scale_fill_brewer(palette = "Set2", name = "Region", drop = FALSE) +
  labs(title = "Characters per skeletal element", x = NULL, y = "Number of characters") +
  theme_minimal(base_size = 14)
show_plot(p_q6, "q6_skeletal_elements", height = 6)
```

Temporal bone and mandible together account for `r percent(sum(hc$skeletal_element %in% c("Temporal", "Mandible")) / n_chars, accuracy = 1)` of all characters. Sampling is uneven, which matters for delimitation because some anatomy carries more weight than other anatomy.

# Q7. Are characters designated as phylogenetic or taxonomic (and are any diagnostic)?

This is directly relevant to species delimitation: it separates characters intended for phylogeny from those used to diagnose taxa.

```{r q7}
q7 <- count_prop(hc, character_designation)
kable(q7, col.names = c("Designation", "n", "Proportion", "%"), digits = 3,
      caption = "Table 7a. Character designation")

q7b <- count_prop(hc, character_significance)
kable(q7b, col.names = c("Significance", "n", "Proportion", "%"), digits = 3,
      caption = "Table 7b. Character significance (only a small subset has been assessed)")

p_q7 <- hc %>% filter(character_designation != "Unscored") %>% droplevels() %>%
  count_prop(anatomical_region, character_designation) %>%
  stack_plot(anatomical_region, character_designation,
             "Character designation by region", "Anatomical region", "Designation", "Dark2")
show_plot(p_q7, "q7_designation_by_region")
```

# Q8. How well is the database linked to anatomical ontology terms (Uberon)?

```{r q8}
uberon_count <- hc %>% count(uberon_ID, sort = TRUE, name = "n_characters")
n_with_id   <- sum(!is.na(hc$uberon_ID))
n_unique_id <- n_distinct(hc$uberon_ID, na.rm = TRUE)
```

`r n_with_id` of `r n_chars` characters (`r percent(n_with_id / n_chars, accuracy = 1)`) have an Uberon ID, covering **`r n_unique_id`** distinct anatomical terms.

```{r q8-top}
p_q8 <- hc %>% filter(!is.na(uberon_ID)) %>%
  count(uberon_ID, skeletal_element, name = "n_characters") %>%
  group_by(uberon_ID) %>% slice_max(n_characters, n = 1, with_ties = FALSE) %>% ungroup() %>%
  slice_max(n_characters, n = 15) %>%
  mutate(label = paste0(skeletal_element, " (", uberon_ID, ")"),
         label = fct_reorder(label, n_characters)) %>%
  ggplot(aes(x = label, y = n_characters)) +
  geom_col(fill = "#FC8D62") +
  coord_flip() +
  labs(title = "15 most frequently scored anatomical structures", x = NULL,
       y = "Number of characters") +
  theme_minimal(base_size = 14)
show_plot(p_q8, "q8_top_uberon_structures", height = 5)

write.csv(uberon_count, "uberon_count.csv", row.names = FALSE)   # same output as the original analysis
```

```{r q8-region}
hc %>%
  group_by(anatomical_region) %>%
  summarise(characters = n(),
            distinct_uberon_terms = n_distinct(uberon_ID, na.rm = TRUE),
            missing_uberon = sum(is.na(uberon_ID)),
            .groups = "drop") %>%
  filter(characters > 0) %>%
  kable(caption = "Table 8. Uberon coverage by anatomical region")
```

# Q9. Where did the characters come from?

The four source columns hold the source's own description of a character, so a non-empty cell means the character appears in that source.

```{r q9, results = "asis"}
src_cols <- unname(source_cols)
if (length(src_cols) == 0) {
  cat("Skipped: none of the source columns were found in the CSV.\n")
} else {

  src_present <- hc %>% select(all_of(src_cols)) %>%
    mutate(across(everything(), ~ !is.na(.x) & str_trim(as.character(.x)) != ""))

  cat("\n\n")
  tibble(Source = src_cols, `Characters` = colSums(src_present)) %>%
    mutate(`% of database` = percent(Characters / n_chars, accuracy = 1)) %>%
    kable(caption = "Table 9a. Characters documented in each source") %>% print()
  cat("\n\n")

  hc %>%
    mutate(`Sources` = apply(src_present, 1, function(r) if (any(r)) paste(src_cols[r], collapse = " + ") else "None")) %>%
    count(Sources, sort = TRUE, name = "n") %>%
    kable(caption = "Table 9b. Which combinations of sources documented each character") %>% print()
  cat("\n\n")
}
```

# Q10. Where are the gaps? (curation checklist)

```{r q10}
issues <- bind_rows(
  hc %>% filter(anatomical_region == "Unscored")      %>% transmute(id, character, issue = "No anatomical region"),
  hc %>% filter(character_type == "Unscored")         %>% transmute(id, character, issue = "No character type"),
  hc %>% filter(multistate_type == "Unscored")        %>% transmute(id, character, issue = "No ordered/unordered designation"),
  hc %>% filter(character_designation == "Unscored")  %>% transmute(id, character, issue = "No phylogenetic/taxonomic designation"),
  hc %>% filter(is.na(uberon_ID))                     %>% transmute(id, character, issue = "No Uberon ID"),
  hc %>% filter(skeletal_element == "Unscored")       %>% transmute(id, character, issue = "No skeletal element"),
  hc %>% filter(is.na(number_of_states))              %>% transmute(id, character, issue = "No number of states"),
  hc %>% filter(number_of_states == 1)                %>% transmute(id, character, issue = "Only 1 state recorded (check)")
)

issues %>% count(issue, sort = TRUE, name = "n_characters") %>%
  kable(caption = "Table 10a. Characters needing attention, by issue")

# Full list to work from in the spreadsheet
write.csv(issues %>% arrange(id), "character_curation_checklist.csv", row.names = FALSE)
```

# Q11. Are "qualitative" characters really measurements?

Only `r sum(hc$character_type == "Continuous")` characters are flagged continuous, but many discrete characters are named like measurements (ratios, widths, heights). This is a keyword screen of the character *names*, so read the flagged list before drawing conclusions. Edit the pattern to tune it.

```{r q11}
hc <- hc %>% mutate(measurement_like = str_detect(character, regex(measure_pattern, ignore_case = TRUE)))

q11 <- hc %>%
  filter(character_type %in% c("Continuous", "Discrete")) %>% droplevels() %>%
  count(character_type, measurement_like) %>%
  mutate(measurement_like = if_else(measurement_like, "Name suggests a measurement", "Name does not"))
kable(pivot_wider(q11, names_from = measurement_like, values_from = n, values_fill = 0),
      caption = "Table 11. Character type vs. measurement-like names")

# Discrete characters whose names suggest a measurement (candidates for binned quantitative traits)
q11_review <- hc %>% filter(character_type == "Discrete", measurement_like) %>%
  select(id, character, skeletal_element, number_of_states)
write.csv(q11_review, "measurement_like_discrete_characters.csv", row.names = FALSE)
```

`r nrow(q11_review)` discrete characters have measurement-like names (full list saved to `measurement_like_discrete_characters.csv`). If most of these really are binned measurements, the true quantitative share of the database is higher than the continuous/discrete flag alone suggests.

# Q12. Which taxa do the characters emphasize?

Counts the characters whose description text (the `description` field plus the Dembo and Strait source text) mentions each taxon. This shows which taxa the character sources were written around, not how each taxon was scored.

```{r q12, fig.height = 5.5}
taxa_tbl <- tribble(
  ~taxon,              ~pattern,                    ~group,
  "P. robustus",       "robustus",                  "Paranthropus",
  "P. boisei",         "boisei",                    "Paranthropus",
  "P. aethiopicus",    "aethiopicus",               "Paranthropus",
  "A. africanus",      "africanus",                 "Australopithecus",
  "A. afarensis",      "afarensis",                 "Australopithecus",
  "A. anamensis",      "anamensis",                 "Australopithecus",
  "A. garhi",          "garhi",                     "Australopithecus",
  "A. sediba",         "sediba",                    "Australopithecus",
  "Ar. ramidus",       "ramidus",                   "Other early hominins",
  "O. tugenensis",     "tugenensis",                "Other early hominins",
  "S. tchadensis",     "tchadensis",                "Other early hominins",
  "H. habilis",        "habilis",                   "Homo",
  "H. rudolfensis",    "rudolfensis",               "Homo",
  "H. ergaster",       "ergaster",                  "Homo",
  "H. erectus",        "erectus",                   "Homo",
  "H. naledi",         "naledi",                    "Homo",
  "H. sapiens",        "sapiens",                   "Homo",
  "Pan",               "\\bPan\\b",                 "Extant apes",
  "Gorilla",           "Gorilla",                   "Extant apes",
  "Pongo",             "Pongo",                     "Extant apes"
)

text_cols <- intersect(c("description", unname(source_cols)), names(hc))
all_text <- hc %>% select(all_of(text_cols)) %>%
  mutate(across(everything(), ~ replace_na(as.character(.x), ""))) %>%
  unite("txt", everything(), sep = " ") %>% pull(txt)

q12 <- taxa_tbl %>%
  mutate(n_characters = map_int(pattern, ~ sum(str_detect(all_text, regex(.x, ignore_case = !str_detect(.x, "Pan"))))))

kable(q12 %>% select(taxon, group, n_characters) %>% arrange(desc(n_characters)),
      caption = "Table 12. Characters whose text mentions each taxon")

p_q12 <- q12 %>% mutate(taxon = fct_reorder(taxon, n_characters)) %>%
  ggplot(aes(x = taxon, y = n_characters, fill = group)) +
  geom_col() + coord_flip() +
  scale_fill_brewer(palette = "Dark2", name = NULL) +
  labs(title = "Characters mentioning each taxon", x = NULL, y = "Number of characters") +
  theme_minimal(base_size = 14)
show_plot(p_q12, "q12_taxon_mentions", height = 5.5)
```

Taxa with few or no mentions (`r paste(q12$taxon[q12$n_characters == 0], collapse = ", ")`) are poorly represented in the character descriptions, which matters if they are among the taxa you want to delimit.

# Q13. Do the state counts agree with the listed states?

`state_code` lists each state as `Label (n)`. Comparing the number of distinct states listed there with `number_of_states` is a quick internal-consistency check. Most mismatches are not parsing noise: for example, `state_code` lists two states ("Absent (0); Present (1)") while `number_of_states` says 3, or lists three while the count says 2. Treat this as a review list.

```{r q13}
parse_states <- function(s) {
  if (is.na(s) || str_trim(s) == "") return(NA_integer_)
  length(unique(str_extract_all(s, "\\((\\d+)\\)")[[1]]))
}

q13 <- hc %>%
  mutate(states_listed = map_int(state_code, parse_states),
         diff = states_listed - number_of_states) %>%
  filter(!is.na(diff))

n_compare  <- nrow(q13)
n_mismatch <- sum(q13$diff != 0)

q13 %>% count(diff, name = "n_characters") %>%
  mutate(meaning = case_when(diff == 0 ~ "Agree",
                             diff > 0  ~ "More states listed than recorded",
                             TRUE      ~ "Fewer states listed than recorded")) %>%
  kable(caption = "Table 13. Listed states minus recorded number_of_states")

write.csv(q13 %>% filter(diff != 0) %>% select(id, character, number_of_states, states_listed, state_code),
          "state_count_mismatches.csv", row.names = FALSE)
```

Of `r n_compare` characters with both fields, `r n_mismatch` (`r percent(n_mismatch / n_compare, accuracy = 1)`) disagree. The list is saved to `state_count_mismatches.csv`. Resolve these before using state counts as simulation parameters (which is what the parameter section below does).

# Q14. Does the ordered/unordered column match the source text?

The Dembo et al. (2015) description text states "(Ordered)" or "(Unordered)". This compares that statement with the `multistate_type` column.

```{r q14, results = "asis"}
if (is.na(dembo_col)) {
  cat("Skipped: no Dembo column was found in the CSV.\n")
} else {
  hc <- hc %>%
    mutate(dembo_order = case_when(
      str_detect(.data[[dembo_col]], regex("\\(Unordered\\)", ignore_case = TRUE)) ~ "Unordered",
      str_detect(.data[[dembo_col]], regex("\\(Ordered\\)",   ignore_case = TRUE)) ~ "Ordered",
      TRUE ~ NA_character_))

  q14 <- hc %>% filter(!is.na(dembo_order), multistate_type != "Unscored") %>%
    mutate(dembo_order = factor(dembo_order, levels = c("Ordered", "Unordered")))

  cat("\n\n")
  print(kable(add_totals(table(`Dembo text` = q14$dembo_order, `multistate_type column` = q14$multistate_type)),
              caption = "Table 14. Ordering stated in Dembo text vs. multistate_type column"))
  cat("\n\n")

  n_agree <- sum(as.character(q14$dembo_order) == as.character(q14$multistate_type))

  write.csv(q14 %>% filter(as.character(dembo_order) != as.character(multistate_type)) %>%
              select(id, character, multistate_type, dembo_order),
            "ordering_disagreements.csv", row.names = FALSE)
}
```

`r if (exists("n_agree")) paste0(n_agree, " of ", nrow(q14), " comparable characters (", percent(n_agree / nrow(q14), accuracy = 1), ") agree.") else "Skipped: no Dembo column was found in the CSV (see the columns listed in the load section)."` Any disagreements are saved to `ordering_disagreements.csv` (empty if the two fields agree everywhere).

# Q15. Is region assignment consistent within a skeletal element?

Some elements appear under more than one region. That may be intentional (for example, the zygomatic contributes to the face), but the rule should be explicit.

```{r q15}
q15 <- hc %>% filter(anatomical_region != "Unscored", skeletal_element != "Unscored") %>%
  count(skeletal_element, anatomical_region, name = "n") %>%
  group_by(skeletal_element) %>% filter(n_distinct(anatomical_region) > 1) %>% ungroup()

q15 %>%
  pivot_wider(names_from = anatomical_region, values_from = n, values_fill = 0) %>%
  kable(caption = "Table 15. Skeletal elements spanning more than one region")

# Characters sitting in the minority region for their element (candidates to double-check)
major <- q15 %>% group_by(skeletal_element) %>% slice_max(n, n = 1, with_ties = FALSE) %>%
  ungroup() %>% select(skeletal_element, major_region = anatomical_region)

q15_review <- hc %>%
  inner_join(major, by = "skeletal_element") %>%
  filter(anatomical_region != major_region, anatomical_region != "Unscored") %>%
  transmute(id, character, skeletal_element, assigned_region = anatomical_region, majority_region = major_region)
write.csv(q15_review, "region_assignment_review.csv", row.names = FALSE)

p_q15 <- q15 %>% stack_plot(skeletal_element, anatomical_region, "Elements split across regions",
                            "Skeletal element", "Region")
show_plot(p_q15, "q15_elements_split_across_regions")
```

`r nrow(q15_review)` characters sit in a non-majority region for their element (saved to `region_assignment_review.csv`).

# Q16. How many characters allow a variable (polymorphic) state?

Fossil samples are variable within species, so how polymorphism is handled matters for delimitation. This counts characters whose `state_code` or source text (Dembo, Strait) describes a "variable" or "polymorphic" state. Source text such as "(3) variable" means the source scored variable taxa as their own state.

```{r q16}
var_pattern <- "variable|polymorph"
src_text_cols <- unname(source_cols)
source_text <- if (length(src_text_cols) == 0) rep("", n_chars) else
  hc %>% select(all_of(src_text_cols)) %>%
    mutate(across(everything(), ~ replace_na(as.character(.x), ""))) %>%
    unite("txt", everything(), sep = " ") %>% pull(txt)

q16 <- hc %>%
  mutate(variable_in_states = str_detect(replace_na(state_code, ""), regex(var_pattern, ignore_case = TRUE)),
         variable_in_text   = str_detect(source_text, regex("\\(\\d+\\)\\s*variable", ignore_case = TRUE)),
         any_variable       = variable_in_states | variable_in_text)

tibble(Field = c("state_code lists a variable/polymorphic state",
                 "Source text defines a numbered 'variable' state",
                 "Either"),
       Characters = c(sum(q16$variable_in_states), sum(q16$variable_in_text), sum(q16$any_variable))) %>%
  mutate(`% of database` = percent(Characters / n_chars, accuracy = 0.1)) %>%
  kable(caption = "Table 16. Characters that allow a variable state")

q16 %>% filter(any_variable) %>% select(id, character, state_code) %>% kable()
```

# Parameterizing simulations from the real database

These are the empirical proportions a simulation (such as MorphoDelim) would need so that simulated character sets mirror the real database. Values are computed only from characters that have been scored for the relevant field.

```{r params}
scored <- function(x) x[!x %in% c("Unscored", "Not assessed")]
prop_of <- function(x, level) mean(scored(as.character(x)) == level)

params <- tribble(
  ~parameter,                         ~value,
  "n_characters",                     n_chars,
  "prop_continuous",                  prop_of(hc$character_type, "Continuous"),
  "prop_discrete",                    prop_of(hc$character_type, "Discrete"),
  "prop_ordered",                     prop_of(hc$multistate_type, "Ordered"),
  "prop_unordered",                   prop_of(hc$multistate_type, "Unordered"),
  "prop_cranial",                     prop_of(hc$anatomical_region, "Cranial"),
  "prop_dental",                      prop_of(hc$anatomical_region, "Dental"),
  "prop_craniofacial",                prop_of(hc$anatomical_region, "Craniofacial"),
  "prop_postcranial",                 prop_of(hc$anatomical_region, "Postcranial"),
  "prop_phylogenetic_designation",    prop_of(hc$character_designation, "Phylogenetic"),
  "prop_taxonomic_designation",       prop_of(hc$character_designation, "Taxonomic"),
  "prop_both_designation",            prop_of(hc$character_designation, "Both"),
  "mean_number_of_states",            mean(hc$number_of_states, na.rm = TRUE),
  "prop_binary",                      mean(hc$number_of_states == 2, na.rm = TRUE)
)
kable(params %>% mutate(value = round(value, 3)), caption = "Table 17. Empirical parameters")

state_dist <- hc %>% filter(!is.na(number_of_states)) %>%
  count(number_of_states, name = "n") %>% mutate(prop = round(n / sum(n), 3))
kable(state_dist, caption = "Table 18. Distribution of number of states (for drawing simulated state counts)")

write.csv(params, "morphodelim_parameters.csv", row.names = FALSE)
write.csv(state_dist, "morphodelim_state_count_distribution.csv", row.names = FALSE)
```

Optional: if you have a simulated character table with a `number_of_states` column, compare it with the real distribution (uncomment and edit).

```{r params-compare, eval=FALSE}
# sim_chars <- read.csv("simulated_characters.csv")            # your MorphoDelim output
# bind_rows(
#   state_dist %>% transmute(number_of_states, prop, source = "Real database"),
#   sim_chars %>% count(number_of_states) %>% mutate(prop = n / sum(n), source = "Simulated") %>%
#     select(number_of_states, prop, source)
# ) %>%
#   ggplot(aes(factor(number_of_states), prop, fill = source)) +
#   geom_col(position = "dodge") +
#   labs(x = "Number of states", y = "Proportion of characters") + theme_minimal(base_size = 14)
```

# Analyses that need a taxon-by-character matrix

The database currently describes characters, not how each taxon was scored. The analyses below need a **taxon x character matrix**, and they run automatically once the file exists.

**Expected file:** `Hominin_Character_Matrix.csv` in the same folder, with one row per taxon (or specimen/OTU) and one column per character.

- The first column is the taxon name; every other column header contains the character `id` from this database (for example `1`, `2`, ... or `char_1`, `char_2`, ...).
- Cells hold the state number (`0`, `1`, `2`, ...). Missing data is `?` or blank, and inapplicable is `-`.
- A polymorphic score lists both states, for example `0/1` or `01`.

```{r matrix-load, results = "asis"}
#matrix_file <- "Hominin_Character_Matrix.csv"
#have_matrix <- file.exists(matrix_file)
#if (!have_matrix) {
 # cat("\n> **No matrix found** (`", matrix_file, "`), so the sections below are skipped. ",
     # "Add the file next to this script and re-knit.\n", sep = "")
#}
```

```{r matrix-prep, eval = have_matrix}
#mat_raw <- readr::read_csv(matrix_file, col_types = readr::cols(.default = "c"),
               #            name_repair = "minimal", show_col_types = FALSE)

#to_set <- function(x) {
 # x <- str_trim(x)
 # if (is.na(x) || x %in% c("", "?", "-", "NA", "N/A")) return(character(0))
 # str_extract_all(x, "[0-9]")[[1]]
#}

#mat_long <- mat_raw %>%
 # rename(taxon = 1) %>%
 # pivot_longer(-taxon, names_to = "char_col", values_to = "raw") %>%
 # mutate(id  = as.integer(str_extract(char_col, "\\d+")),
         #set = map(raw, to_set),
         #scorable = lengths(set) > 0,
         #polymorphic = lengths(set) > 1) %>%
 # filter(!is.na(id)) %>%
  #left_join(hc %>% select(id, anatomical_region, character_type, number_of_states, character), by = "id")

#n_taxa <- n_distinct(mat_long$taxon)

# Characters in the matrix but not in the database (check for mistyped ids)
#setdiff(unique(mat_long$id), hc$id)
```

## M1. How many characters can be scored for each taxon?

```{r m1, eval = have_matrix, fig.height = 5}
#m1 <- mat_long %>%
 # group_by(taxon) %>%
  #summarise(scored = sum(scorable), polymorphic = sum(polymorphic),
   #         pct_of_database = scored / n_chars, .groups = "drop") %>%
  #arrange(desc(scored))
#kable(m1 %>% mutate(pct_of_database = percent(pct_of_database, accuracy = 1)),
 #     caption = "Table M1. Characters scored per taxon")

#p_m1a <- ggplot(m1, aes(x = fct_reorder(taxon, scored), y = pct_of_database)) +
 # geom_col(fill = "#4C78A8") + coord_flip() +
  #scale_y_continuous(labels = percent, limits = c(0, 1)) +
  #labs(title = "Share of database characters scored per taxon", x = NULL, y = "Characters scored") +
  #theme_minimal(base_size = 14)
#show_plot(p_m1a, "m1_scored_per_taxon", height = 5)

# By region: does missing data fall on particular parts of the skeleton?
#p_m1b <- mat_long %>% filter(anatomical_region %in% c("Cranial", "Dental", "Craniofacial")) %>%
 # group_by(taxon, anatomical_region) %>%
  #summarise(pct_scored = mean(scorable), .groups = "drop") %>%
  #ggplot(aes(x = anatomical_region, y = taxon, fill = pct_scored)) +
  #geom_tile(colour = "white") +
  #geom_text(aes(label = percent(pct_scored, accuracy = 1)), size = 3) +
  #scale_fill_gradient(low = "#F7F7F7", high = "#4C78A8", labels = percent, name = "Scored") +
  #labs(title = "Completeness by taxon and region", x = NULL, y = NULL) +
  #theme_minimal(base_size = 13)
#show_plot(p_m1b, "m1_completeness_by_region", height = 5)
```

## M2. Missing-data pattern across characters

```{r m2, eval = have_matrix, fig.height = 6}
#m2 <- mat_long %>%
 # group_by(id, anatomical_region) %>%
  #summarise(taxa_scored = sum(scorable), pct_scored = mean(scorable), .groups = "drop")

#p_m2a <- ggplot(m2, aes(x = pct_scored)) +
 # geom_histogram(binwidth = 0.1, boundary = 0, fill = "#66C2A5", colour = "white") +
  #scale_x_continuous(labels = percent) +
  #labs(title = "How many taxa can each character be scored for?", x = "Share of taxa scored", y = "Number of characters") +
  #theme_minimal(base_size = 14)
#show_plot(p_m2a, "m2_taxa_scored_per_character", height = 4.5)

# Character x taxon completeness map, characters ordered by region then id
#p_m2b <- mat_long %>%
 # mutate(id = factor(id, levels = hc %>% arrange(anatomical_region, id) %>% pull(id)),
  #       taxon = factor(taxon, levels = m1$taxon)) %>%
  #ggplot(aes(x = id, y = taxon, fill = scorable)) +
  #geom_tile() +
  #scale_fill_manual(values = c(`TRUE` = "#4C78A8", `FALSE` = "#EEEEEE"), labels = c("Missing", "Scored"), name = NULL) +
  #labs(title = "Scoring map (characters ordered by region)", x = "Characters", y = NULL) +
  #theme_minimal(base_size = 13) + theme(axis.text.x = element_blank(), panel.grid = element_blank())
#show_plot(p_m2b, "m2_scoring_map", height = 5)

#n_rarely_scored <- sum(m2$pct_scored < 0.5)
```

`r if (exists("n_rarely_scored")) paste(n_rarely_scored, "characters are scored for fewer than half of the taxa.")`

## M3. Which characters actually vary across taxa?

An invariant character cannot separate taxa. Variability here is the number of distinct states observed across scored taxa.

```{r m3, eval = have_matrix}
#m3 <- mat_long %>% filter(scorable) %>%
  #group_by(id, anatomical_region, number_of_states) %>%
  #summarise(taxa_scored = n(), states_observed = n_distinct(unlist(set)),
   #         any_polymorphic = any(polymorphic), .groups = "drop") %>%
  #mutate(class = case_when(states_observed <= 1 ~ "Invariant", TRUE ~ "Variable"))

#m3 %>% count(class, name = "n_characters") %>% mutate(pct = percent(n_characters / sum(n_characters), accuracy = 1)) %>%
 # kable(caption = "Table M3a. Invariant vs. variable characters across the sampled taxa")

#p_m3 <- m3 %>% filter(anatomical_region %in% c("Cranial", "Dental", "Craniofacial")) %>%
 # count(anatomical_region, class, name = "n") %>%
  #mutate(anatomical_region = droplevels(anatomical_region), class = factor(class)) %>%
  #stack_plot(anatomical_region, class, "Variable characters by region", "Region", "Class", "Set2")
#show_plot(p_m3, "m3_variable_characters_by_region")

# Are all the states listed in the database actually observed?
#kable(table(`Recorded number_of_states` = m3$number_of_states, `States observed in matrix` = m3$states_observed),
 #     caption = "Table M3b. Recorded vs. observed number of states")
```

## M4. Which characters separate which taxa? (pairwise fixed differences)

A character is counted as separating two taxa when both are scored and their state sets do not overlap (no shared state, polymorphism included). This is a descriptive screen, not a species-delimitation test.

```{r m4, eval = have_matrix, fig.height = 6}
#taxa <- sort(unique(mat_long$taxon))
#ids  <- sort(unique(mat_long$id))

#S <- matrix(list(character(0)), nrow = length(taxa), ncol = length(ids),
 #           dimnames = list(taxa, as.character(ids)))
#for (k in seq_len(nrow(mat_long))) {
 # S[[mat_long$taxon[k], as.character(mat_long$id[k])]] <- mat_long$set[[k]]
#}

#pairs <- t(combn(taxa, 2))
#pair_res <- map_dfr(seq_len(nrow(pairs)), function(k) {
 # a <- S[pairs[k, 1], ]; b <- S[pairs[k, 2], ]
  #comparable <- lengths(a) > 0 & lengths(b) > 0
  #fixed <- map2_lgl(a, b, ~ length(.x) > 0 && length(.y) > 0 && length(intersect(.x, .y)) == 0)
  #tibble(taxon_a = pairs[k, 1], taxon_b = pairs[k, 2], id = ids, comparable = comparable, fixed = fixed)
#}) %>% left_join(hc %>% select(id, anatomical_region, character), by = "id")

#pair_summary <- pair_res %>%
 # group_by(taxon_a, taxon_b) %>%
  #summarise(comparable = sum(comparable), fixed_differences = sum(fixed), .groups = "drop")

#p_m4 <- pair_summary %>%
 # bind_rows(pair_summary %>% rename(taxon_a = taxon_b, taxon_b = taxon_a)) %>%
  #ggplot(aes(x = taxon_a, y = taxon_b, fill = fixed_differences)) +
  #geom_tile(colour = "white") +
  #geom_text(aes(label = fixed_differences), size = 3) +
  #scale_fill_gradient(low = "#F7F7F7", high = "#FC8D62", name = "Separating\ncharacters") +
  #labs(title = "Characters separating each pair of taxa", x = NULL, y = NULL) +
  #theme_minimal(base_size = 12) + theme(axis.text.x = element_text(angle = 45, hjust = 1))
#show_plot(p_m4, "m4_pairwise_separating_characters", height = 6)

# Pairs that are hardest to tell apart, and pairs with too little shared data to judge
#kable(pair_summary %>% arrange(fixed_differences) %>% head(10),
 #     caption = "Table M4a. Ten pairs with the fewest separating characters")
#kable(pair_summary %>% arrange(comparable) %>% head(10),
 #     caption = "Table M4b. Ten pairs with the least comparable data")

# Which regions supply the separating characters?
#pair_res %>% filter(fixed, anatomical_region %in% c("Cranial", "Dental", "Craniofacial")) %>%
 # count(anatomical_region, name = "n") %>% mutate(pct = percent(n / sum(n), accuracy = 1)) %>%
  #kable(caption = "Table M4c. Region of separating characters (all pairs)")

# Characters that separate the most pairs
#pair_res %>% filter(fixed) %>% count(id, character, anatomical_region, name = "pairs_separated") %>%
 # slice_max(pairs_separated, n = 15) %>%
  #kable(caption = "Table M4d. Characters that separate the most taxon pairs")
```

# Summary slide table

One table you can paste directly into a slide.

```{r summary}
bind_rows(
  q1 %>% transmute(Question = "Character type",    Category = as.character(character_type),        n, pct),
  q2 %>% transmute(Question = "Anatomical region", Category = as.character(anatomical_region),     n, pct),
  q4 %>% transmute(Question = "Ordered vs. unordered", Category = as.character(multistate_type),   n, pct),
  q7 %>% transmute(Question = "Designation",       Category = as.character(character_designation), n, pct)
) %>%
  kable(col.names = c("Question", "Category", "n", "%"))
```

# Save all figures

```{r save-all-figures}
# One multi-page PDF containing every figure in report order (landscape, ready to print)
combined <- file.path(fig_dir, "all_figures.pdf")
if (length(.figs$plots) > 0) {
  pdf(combined, width = 11, height = 8.5)
  for (p in .figs$plots) print(p)
  invisible(dev.off())
  cat("Saved", length(.figs$plots), "figures to", normalizePath(fig_dir),
      "as", paste(fig_formats, collapse = " + "), "and combined them in", basename(combined), "\n")
  print(sort(list.files(fig_dir)))
}
```
