---
title: "HC Analysis"
author: "Jyhreh Johnson"
date: "2025-09-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE) # Toggle whether to show R code chunks
```

```{r echo = FALSE, include=FALSE}
# Preliminaries, Load necessary libraries
library(ggplot2)
library(tidyverse)
library(dplyr)
library(jsonlite)
library(ggstatsplot)
library(knitr)
```

# Load/Reload Data 
```{Load Data}
#Load Hominin Characters spreadsheet
hominin_characters <- read.csv("Known_Hominin_Characters.csv") #Make sure to check spreadsheet in Google Drive for changes


# Load/Reload Data from Origins using the API [Not working yet]
#omo_turkana_URL <- "http://paleocore.org/origins/api/turkfossils/"
#omo_turkana_df <- as.data.frame(fromJSON(omo_turkana_URL))
```

### Character Type Count
```{r table.cap='Table 1'}
chartype.table <-hominin_characters %>% count(character_type) # create summary df
chartype.table.cd<-chartype.table[chartype.table$character_type %in% c('Continuous','Discrete'),] # subset df to show just Continuous/Discrete rows
chartype.table.cd$proportions <- chartype.table.cd$n/sum(chartype.table.cd$n) # Add column with proportion data
knitr::kable(chartype.table, label=NA, caption='Table 1. Chartacter Type counts')

chartype_plot <- barplot(chartype.table$n, 
                                  names.arg = chartype.table$character_type, 
                                  col = "pink",
                                  main = "Character Type Plot", 
                                  xlab = "Character Type", 
                                  ylab = "Count")

text(x = chartype_plot, 
     y = chartype.table$n / 2,  # Middle of bars
     labels = chartype.table$n, 
     col = 'black',
     font = 2)  # Text size
```


### Multistate Type Count
```{r table.cap='Table 2'}
multitype.table <-hominin_characters %>% count(multistate_type) # create summary df
multitype.table$proportions <- multitype.table$n/sum(multitype.table$n) # Add column with proportion data
knitr::kable(chartype.table, label=NA, caption='Table 2. Multistate Type counts')

multitype_plot <- barplot(multitype.table$n, 
                                  names.arg = multitype.table$multistate_type, 
                                  col = "lightblue",
                                  main = "Multistate Type Plot", 
                                  xlab = "Multistate Type", 
                                  ylab = "Count")

text(x = multitype_plot, 
     y = multitype.table$n / 2,  # Middle of bars
     labels = multitype.table$n, 
     col = 'black',
     font = 2)  # Text size
```


### Anatomical Region Count
```{r table.cap='Table 3'}
anatomical_region.table <-hominin_characters %>% count(anatomical_region) # create summary df
anatomical_region.table$proportions <- anatomical_region.table$n/sum(anatomical_region.table$n) # Add column with proportion data
knitr::kable(chartype.table, label=NA, caption='Table 3. Anatomical counts')

anatomical_region_plot <- barplot(anatomical_region.table$n, 
                                  names.arg = anatomical_region.table$anatomical_region, 
                                  col = "lightgreen",
                                  main = "Anatomical Region Plot", 
                                  xlab = "Anatomical Region", 
                                  ylab = "Count")

text(x = anatomical_region_plot, 
     y = anatomical_region.table$n / 2,  # Middle of bars
     labels = anatomical_region.table$n, 
     col = 'black',
     font = 2)  # Text size
```

### Percentage Breakdown of Anatomical Region x Character Type
https://indrajeetpatil.github.io/ggstatsplot/articles/web_only/ggbarstats.html
```{BarStackPlot}
#BarStackPlot
chartype_bd_plot <- ggbarstats(data = hominin_characters, 
                               x = anatomical_region, 
                               y = character_type,
                               title = "Anatomical Region by Character Type") #show the anatomical region stats for each character type
chartype_bd_plot
```
### Statistical Information
The p-values (p = 0.39, p = 3.13e-07, p = 6.47e-85) test whether the distribution of traits across regions differs significantly from expected
Very small p-values indicate statistically significant differences in how traits are distributed across anatomical regions
The bottom statistics show effect sizes and confidence intervals

### Interpretation
This suggests that:
Cranial region dominates both continuous and discrete traits (~64-65%)
Dental traits make up a substantial portion, especially for continuous measurements (30% vs 26%)
Craniofacial traits are the least common in both categories
The distribution patterns are statistically significantly different from random, indicating systematic differences in how different types of traits are distributed across cranial anatomy

### Percentage Breakdown of Anatomical Region x Multistate Type
```{BarStackPlot}
#BarStackPlot
multitype_bd_plot <- ggbarstats(data = hominin_characters, 
                               x = anatomical_region, 
                               y = multistate_type,
                               title = "Anatomical Region by Multistate Type") #show the anatomical region stats for each character type
multitype_bd_plot
```

###Statistical Significance
The p-values show significant differences in trait distribution
Very small p-values (1.39e-22, 9.09e-70) indicate the distributions are highly statistically significant
This suggests that ordered vs. unordered traits are not randomly distributed across anatomical regions

###Interpretation
Cranial region dominates both ordered and unordered traits (~64-66%)
Dental traits are more common in ordered categories, which makes sense since many dental features (like wear, eruption stages) follow natural progressions
Craniofacial traits remain the least common across all categories
The similar distributions between ordered and unordered traits suggest consistent patterns in how cranial anatomy is characterized, regardless of whether traits follow natural sequences

```{Uberon ID Analysis}
uberon_count <- hominin_characters %>% count(uberon_ID) #count number of Uberon ID occurrences
write.csv(uberon_count, "uberon_count.csv")
```